In-silico molecular enrichment and clearance of the human intracranial space

### 2025-01-23 paper/main.tex
Shuffle parts of the Discussion.

\newcommand{\rami}[1]{\textcolor{blue}{#1}}
\newcommand{\mer}[1]{\textcolor{magenta}{#1}}
\newcommand{\mar}[1]{\textcolor{violet}{#1}}
\newcommand{\discuss}[1]{\textcolor{red}{#1}}
\newcommand{\fixme}[1]{\textcolor{red}{#1}}
\newcommand{\draft}[1]{\textcolor{gray}{#1}}
%\newcommand{\brain}{\Omega_{\rm{PAR}}} 
%\newcommand{\sas}{\Omega_{\rm CSF}}
%\newcommand{\spinal}{\Gamma_{\rm SAS}}
%\newcommand{\skull}{\Gamma_{\rm{skull}}}
%\newcommand{\gin}{g_{\rm{influx}}}

@@

\begin{abstract}
``The abstract — which should be no more than 200 words long and contain
no references — should serve both as a general introduction to the
topic and as a brief, non-technical summary of the main results and
their implications.'' 
\end{abstract}
\begin{document}

@@

\mer{MER: Target length: 5000 words (Intro, Results, Discussion) plus <3000 words Methods: 700 words Introduction. 2800 words Results. 1500 words Discussion.}

@@

\section*{Introduction (rough ideas)}

% What is the problem? (Big picture)
% - Big picture/importance sentence.
% - More in-depth info
% - "XXX is unknown".

Transport of molecules within the spaces of the central nervous system
(CSF spaces, parenchyma) is important. Both in the context of
intracranial drug delivery with potential for personalized treatment
of cancer, diagnostics and treatment of CSF-related disorders, and in
the context of understanding clearance in connection with
neurodegenerative disease and for future treatment and diagnostics of
dementia disorders. Understanding and predicting molecular enrichment
is important for treatment and diagnostics of (brain) cancer, for
diagnostics of CSF-related disorders like normal pressure
hydrocephalus, subarachnoid hemorrage. Moreover, altered transport and
clearance is believed to be key factor in neurodegenerative diseases
like Alzheimer's disease and Parkinson's disease, and in sleep. CSF
flow and PVS established to play a key role. Clinical studies have
revealed large individual variation and variation between
cohorts/disease groups, therefore important in particular in the
context of personalized medicine. However, we are far from
understanding (the mechanisms behind) these variations.

% What we already know about the problem
Our current understanding has primarily been established via
experimental medical research (cats, rats, mice, pigs). Big open
questions and controversy remain, lack holistic understanding of these
physiological process remains, hard to translate experimental studies
to humans. In the last decade, clinical studies. Clinical studies have
revealed large individual variation and variation between
cohorts/disease groups. However, we are far from understanding the
mechanisms behind these variations and integrating and translating
experimental observations to humans.  In-silico modelling is starting
to play an integral role, evaluating hypothesis and pinpointing
mechanisms, but most efforts have been focused on local aspects and
lacking a holistic model. Hard to quantify the overall contribution of
individual effects.

PVS flow is important as shown by experimental studies. This is being
corroborated by clinical studies in humans (Eide and Ringstad (2024),
and the new one). However, a number of open questions remain: to what
extent do the PVS constitute an ensheathed space, how fast is the flow
in human PVS, what is the effect of dilated PVS as seen in disease
cases. Also, open questions on transport characteristics of the CSF
spaces - what gives the individual variations in tracer enhancements?

What is known is that fluid motion (in the ventricular system and SAS,
in perivascular spaces, and possibly within the parenchyma) and
pulsatility play key
roles. Glymphatics~\cite{iliff2012paravascular}. Diffusion, PVS flow
and dispersion, CSF flow. Molecular transport in perivascular spaces
(PVSs) is established as a key pathway for human brain clearance and
delivery. Previous studies indicate that molecules move rapidly in the
subarachnoid space (SAS) and in PVSs surrounding pial arteries. CSF
production in the choroid plexus and possibly in the parenchyma. CSF
efflux pathways, meningeal lympahtics, nerve routes. See also to
Cserr, Phys Rev., 1971, bulk convective flow may be more important for
larger species (larger distances) and larger molecules. Contribution
both from net flow (convection) and pulsatility (mixing) both in the
CSF spaces and in PVSs. Some aspects are better established, such as;
while other factors are more disputed such as ... contribution of
vasomotion. Moreover, the role of local effects in the bigger picture
has remained hard to evaluate.

% What are our research questions and how do we answer them
Recent developments on 3D-1D modelling of transport in PVS and
tissue. ``Digital twin for molecular enrichment and clearance in human
intracranial spaces''. Can we model the interplay between
molecular transport in the intracranial compartments, CSF flow and PVS
flow and transport to holistically quantify the contributions in humans?

% Main findings and outlook
Here, we aim to model and study this transport in a full pial
perivascular and/or vascular network embedded in the SAS.

\iffalse
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
\mer{Update: ~2800 words and 7 subsections - around 300-450 words each?}
\begin{enumerate}
\item
  ``A digital twin'' for molecular enrichment and clearance in human intracranial spaces
\item
  CSF flow, the role of production versus no/cardiac/respiratory
  dispersive effects on tracer trajectories, flow bifurcation. With and without tentorium cerebellum.
\item
  ``The role'' of PVS flow speed
\item
  ``The role'' of PVS sheaths and permeability - does functional separation entail structural separation?
\item
  ``The role'' of PVS size (motivated by pathologies)
\item
  Clearance from the CSF spaces - with and without low pressure veins, but keeping the intrathecal source of solute. 
\item
  Clearance from the parenchyma: $c_0 = 1.0$. Different substances?
\end{enumerate}

\begin{figure}
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/Brain-PVS-callouts.png}
     \end{subfigure}
     \hfill
     \caption{a) Illustration of the different intracranial domains
       considered with labels;}
     \label{fig:intracranial_domains}
\end{figure}

\subsection*{High-fidelity in-silico predictions of intracranial solute transport and exchange}

\begin{itemize}
\item
  Present mathematical and computational model and geometries (\Cref{tab:scenarios}). 
\item
  Highlight computational model availability, open source code.
\item
  Provides an interactive in-silico platform for studying intracranial solute transport -- distribute such that anyone can run (possibly try to recruit MinRK if advantageous).
\item
  \mer{MER: Remember to report on how much concentration enters into the brain?}
\end{itemize}

\subsection*{CSF flow and dispersion shifts enrichment patterns}

Flow of CSF through the SAS and the ventricular system is a crucial driver of transport in the human brain environment.
While cardiac- and respiratory induced oscillatory flow increases mixing, the production of CSF in the choroid plexus creates a steady advective flow within the CSF-filled spaces of the cranium \cite{hornkjol2022csf}. 
We construct a 3D model of the SAS and the ventricular system (\Cref{fig:csf}A,B and C) and compute CSF flow fields for three different scenarios: CSF production in the choroid plexus, CSF flow at systolic peak blood inflow and CSF flow in response to free breathing.

% production flow simulation
To simulate the production-driven flow, we impose an inflow of $400\,$ml/day \cite{nilsson1992circadian} across the surface of the lateral ventricle and allow a resistive outflow across the upper convexity of the arachnoid mater, leading to a total pressure drop of $26\,$mPa between the production and outflow sites (\Cref{fig:csf}D). A maximum flow velocity of $1.8\,$mm/s is attained in the aqueduct, while the mean velocity in the SAS is $2.6\,\mu$m/s (\Cref{fig:csf} E).

%The Stokes system is solved with the finite element software $FEniCS$, stored and then subsequently read for all the relevant solute transport simulations. See Figure~\ref{fig:csf} for a streamline visualization of the obtained velocity profile and \Cref{sec:details_numerical_method} for details on the numerical algorithm employed.

% cardiac pulsatility
Assessing the effect of cardiac-driven dispersion, we compute a steady CSF flow field at peak systolic blood inflow. We mimic the resulting reduction of CSF space volume by imposing an inflow of $6\,$ml/s \cite{baledent2014imaging, causemann2022human} and $0.31\,$ml/s \cite{vinje2019respiratory} across the upper arachnoid mater ($\Gamma_{\rm AM-U}$) and the lateral ventricle ($\Gamma_{\rm VL}$) surface, respectively. We allow free outflow into the spinal SAS ($\Gamma_{\rm SSAS}$). In addition, we upscale the obtained pressure field to account for the lack of pulsatile inertial effects in our model (see \ref{sec:dispersion_details}). In this scenario, we observe a pressure drop of $10\,$Pa between lateral ventricles and the spinal SAS and a maximum flow velocity of $19.8\,$cm/s in the aqueduct (\Cref{fig:csf}F and J).
Next, we estimate a spatially varying dispersion enhancement factor $R$ from cardiac-induced pulsatility by combining the computed pressure field with the theory of shear-augmented dispersion (see \Cref{sec:dispersion_details}). Assuming a cardiac frequency of $1\,$Hz, find that cardiac-induced pulsatile CSF flow increases the effective diffusion by up to two orders of magnitude in the aqueduct and around the cisterna magna, but has little effect in most of the SAS ($R < 1$) (\Cref{fig:csf} G and K).

% respiratory pulsatility
Employing the same methodology, we estimate the respiratory contribution to dispersive mixing by computing the CSF flow field driven by an inflow of $0.121\,$ml/s \cite{liu2024using} and $1\,$ml/s \cite{gutierrez2022effect} across the lateral ventricular wall and the arachnoid mater, respectively. This amounts to a respiratory peak flow volume of $1.121\,$ ml/s at the craniocervical junction \cite{gutierrez2022effect}. While the resulting flow velocities are only about one fourth of their cardiac-induced counterparts (see \Cref{fig:csf}H and L), respiratory dispersive mixing reaches a factor of up to 350 due the lower respiratory frequency ($0.25\,$Hz) (\Cref{fig:csf}I and M).

% tracer trajectory of the standard model
Next, we use the computed advective velcity field and dispersion enhancement factors to predict the spreading of $0.5\,$mmol intrathecally injected Gadobutrol in CSF-filled spaces, the parenchymal tissue and PVSs around arteries and veines. One hour post-injection, tracer moves upwards in the SAS frontally of the brainstem,  and quickly reaches supratentorial regions (\Cref{fig:csf}N). Here, it spreads both posteriorly through the longitudinal fissure and anteriorly through the outer SAS, reaching the top of the cerebral cortex after around 12 h. After 24h, the tracer covers most of the brains surface (with exception of some posterior regions) and has penetrated several centimeters into the parenchymal tissue.

% tracer trajectory of model variations

\begin{itemize}
    \item only dispersion model: No production flow. Tracer remains in the subtentorial regions, in cerebellum, brain stem and surrounding CSF. In contrast to other models, it travels up the ventricular system - similar to what is observed in iNPH patients?
    \item Low dispersion model: account for uncertainty (or pathologies with impaired pulsatile motion?) - tracer exclusively spreads in the frontal half of the brain. Bifurcation of CSF production flow field: flow bifurcates around the midbrain in two directions: a) through the quadrigeminal cistern into the longitudinal fissure and b) frontally around the brain through the outer SAS. Only models with "enough" dispersion enable tracer to follow trajectory a).
    \item without tentorium: 
\end{itemize}

\begin{figure}[h!]
\centering 
\includegraphics[width = 0.9\linewidth]{figures/figure2.png}
\caption{
A) Illustration of 3D-domains; 
B) Illustration of relevant boundaries; 
C) schematic illustration of the model, its doamins and boundaries; 
D) CSF flow induced by CSF production; 
E) production-induced CSF velocities in the SAS and third and fourth ventricle (V3 \& V4); 
F) CSF flow induced by peak systolic expansion; 
G) cardiac dispersion enhancement factor;
H) CSF flow induced by respiration; 
I) respiratory dispersion enhancement factor;
J) cardiac-induced flow velocities;
K) cardiac-induced dispersion enhancement factor;
L) respiratory flow velocities;
M) respiratory dispersion enhancement factor;
N) sagittal, coronal and axial planes of tracer concentration after 1, 6, 12 and 24$\,$h for model A. The outer model boundaries is shown in cyan, the pial membrane in purple and the arteries in red.;
O) model X,Y and Z after 12h;
}
\label{fig:csf}
\end{figure}

\subsection*{The significance of perivascular spaces as flow highways}
\begin{itemize}
\item
  Perivascular flow and transport has been identified as a key
  mechanism and potential target for enhancing brain drug delivery and
  metabolic waste clearance -- what are the global (intracranial)
  effects of such intracranial highways? 
\item
  Do solutes move from PVS to the SAS, or both along the PVS and along
  the SAS?
\end{itemize}

\paragraph{PVS Flow velocities}
\begin{itemize}
    \item we consider three different PVS flow fields: PVS flow driven by production field pressure differences, cardiac peristaltic flow and vasomotion peristaltic flow. 
\end{itemize}

\begin{figure}
    \centering
    \includegraphics[height=\textheight]{figures/figure3.png}
    \caption{\mer{MER: Put vasculature network illustration here. Many aspects actually. Marking of segments. Location relative to 3D. Radius etc. Marius will evaluate and suggest. Results: Group segments by logical order. For key arterial segments: average radius and length (can be moved to Supplementary if need be); pressure values; flow velocities; and then histograph with flow velocities for each mesh cell. Make sure to minorly note that ACA refers to a segment of the ACA and not the whole thing e.g.}}
    \label{fig:pvs}
\end{figure}
\Cref{fig:pvs}.

\subsection*{Structural versus functional compartmentalization of perivascular spaces}

\begin{itemize}
\item
  Are perivascular spaces functionally or structurally
compartmentalized? Human and rodent observations indicate that tracers
concentrate in perivascular spaces surrounding the pial and
subarachnoid vasculature. Must these spaces be structural i.e.~defined
by semi-permeable structural
membrane~\cite{zhang1990interrelationships, zhang1992directional,
  mestre2018flow, eide2024functional}, or can such
compartmentalization be a result of regions of enhanced flow or
mixing~\cite{bedussi2017paravascular, vinje2021brain}? See also
references in Background.
\item
  Vary the PVS permeability $\xi$.
\end{itemize}

\subsection*{Enlarged perivascular spaces yield ...}

\begin{itemize}
  \item Take $R_2 = 3 R_1$ instead of $2 R_1$.
\end{itemize}

\subsection*{The perivenous contribution to molecular clearance of CSF tracers}

\begin{itemize}
\item
  Look at clearance phase 12h-24h-48h-72h.
\item
  Look at concentration in the veins
\item
  Look at concentration in the parenchyma
\item
  Consider low-pressure sink for the perivenous spaces.
\end{itemize}

\subsection*{Clearance of intraparenchymal solutes}

\begin{itemize}
\item
  Clearance of a molecular concentration in the parenchyma, different
  molecular sizes - ranging from Gadubutrol to amyloid-$\beta$/$\tau$.
\end{itemize}

\bigskip

\FloatBarrier
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

\paragraph{CSF flow model validation}

Hornkjøl et al \cite{hornkjol2022csf} model CSF flow induced by production in the choroid plexus and report a peak velocity of $8.9\,$mm/s in the aqueduct. Given that similar flow rates were enforced, the deviation from our value of $2.0\,$mm/s is likely due to differences in diameter of the reconstructed aqueduct. Additionally, their peak velocities in the SAS range between 20 - 50 $\mu\,$m/s, which is compatible with our mean velocity of $4.4\,\mu$m/s (\Cref{fig:csf_flow_histo}).

The aqueductal ($0.31\,$ml/s) and craniocervical ($8.31\,$ml/s) flow rates predicted by our model align well with previously reported values of $0.31\,$ml/s peak aqueductal flow rate \cite{vinje2019respiratory} and about $9\,$ml/s peak net blood inflow into the cranium \cite{baledent2014imaging}. Further, Linninger et al \cite{linninger2007cerebrospinal} model cardiac-induced CSF flow and predict a maximum transmantle pressure difference of $10\,$Pa, which is in good agreement with our estimate of $8.5\,$Pa as well as clinically measured mean pressure differences of $11.5\,$Pa between sensors placed subdurally and in the lateral ventricle \cite{vinje2019respiratory}.
% 10.7 cm/s CSF flow at the craniovertebral junction in control \cite{bapuraj2016cerebrospinal}, area difference?

Sharp et al used the theory of dispersive mixing to predict a dispersion enhancement factor of $91.8$ in the (open) spinal SAS, which lies well within the range of our spatially varying estimate (ranging from 0 to 160). Conversely, \cite{ray2021quantitative} report up to 10,000 times greater diffusion in perivascular regions, but their modelling lumps diffusive, dispersive and advective transport in a single enhancement factor and is thus not directly comparable.

\subsection*{Limitations}

>>>>

\newcommand{\rami}[1]{\textcolor{blue}{#1}}
\newcommand{\mer}[1]{\textcolor{magenta}{#1}}
\newcommand{\mar}[1]{\textcolor{violet}{#1}}
\newcommand{\mk}[1]{\textcolor{orange}{#1}}
\newcommand{\discuss}[1]{\textcolor{red}{#1}}
\newcommand{\fixme}[1]{\textcolor{red}{#1}}
\newcommand{\draft}[1]{\textcolor{lightgray}{#1}}
%\newcommand{\brain}{\Omega_{\rm{PAR}}} 
%\newcommand{\sas}{\Omega_{\rm CSF}}
%\newcommand{\spinal}{\Gamma_{\rm SAS}}
%\newcommand{\skull}{\Gamma_{\rm{skull}}}
%\newcommand{\gin}{g_{\rm{influx}}}

@@

%% The problem we are addressing here is that we would like to be able to
%% understand, integrate and quantify the role of different mechanisms -
%% including effects associated with individual variations - on molecular
%% enrichment and clearance of human intracranial spaces. And we think of
%% this as a step on the way to predict tracer enrichment after
%% intrathecal injection in a personalized medicine context. Which is
%% important because of drug delivery/treatment in cancer, diagnostics
%% for CSF and dementia disorders, and clearance in neurodegenerative
%% disease.

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to
  human brain health and function, with alterations often linked to
  disease and functional impairment, and with distinct opportunities
  for personalized diagnostics and treatment. However, our
  understanding of these mechanisms and their interplay remains
  incomplete, in part due to the complexity of integrating insights
  across scales, between species and from different modalities. Here,
  we combine geometric model reduction and mixed-dimensional
  modelling, multi-modal magnetic resonance images, and high
  performance computing to construct and explore a high-fidelity
  in-silico model of human intracranial molecular enrichment and
  clearance. This model predicts the temporal and spatial spreading of
  a solute within an image-derived geometric representation of the
  subarachnoid space, ventricular system and brain parenchyma,
  including branching networks of surface perivascular spaces
  (PVSs). Our findings highlight the significant impact of
  cerebrospinal fluid (CSF) production and intracranial pulsatility on
  molecular enrichment following intrathecal tracer injection. We
  demonstrate that low-frequency vasomotion induces moderate CSF flow
  rates in surface PVS networks that substantially enhance tracer
  enrichment, and that impaired enrichment is a natural direct
  consequence of enlarged PVSs. This openly available technology
  platform thus provides an opportunity for integrating separate
  observations on diffusion in neuropil, vascular dynamics,
  intracranial pulsatility, CSF production, and efflux, and for
  exploring questions relating to drug delivery and clearance in the
  human brain environment.
\end{abstract}
\begin{document}

@@

%\mer{MER: Target length: 5000 words (Intro, Results, Discussion) plus <3000 words Methods: 700 words Introduction. 3100 words Results. 1200 words Discussion.}

\iffalse
\paragraph*{Abbreviations}
\begin{itemize}
\item
  Subarachnoid space (SAS)
\item
  Cerebrospinal fluid (CSF)
\item
  Perivascular space (PVS), perivascular spaces (PVSs)
\end{itemize}
\fi

@@

% What is the big picture problem and why is it important? 
The mechanisms underlying molecular transport within the intracranial
space are fundamental to human brain health and
function. Neurodegenerative diseases such as Alzheimer's, Parkinson's,
and Huntington's disease are all associated with abnormal accumulation
of protein aggregates together with alterations in transport and
clearance characteristics~\cite{rasmussen2018glymphatic,
  harrison2020impaired, eide2023plasma, liu2024glymphatic}. Moreover,
sleep and conversely sleep-deprivation play a definite yet enigmatic
role in modulating molecular enrichment and
clearance~\cite{xie2013sleep, eide2021sleep, eide2022altered,
  miao2024brain, hauglund2025norepinephrine}. In the last decade,
established theories have been challenged by new findings revealing a
greater degree of molecular movement and
exchange~\cite{iliff2012paravascular, ringstad2017glymphatic,
  louveau2017understanding, proulx2021cerebrospinal,
  bohr2022glymphatic}, as well as substantial variability in
enrichment and clearance between individuals and between patient
cohorts~\cite{ringstad2018brain, eide2021direction, eide2021impaired,
  eide2022altered}. These observations provide distinct opportunities
(and challenges) for personalized medicine e.g.~for tailored
intrathecal delivery of chemotherapy~\cite{lohela2022glymphatic} and
for early diagnostics of impaired brain
clearance~\cite{eide2021clinical, van2024human}. In spite of their
importance, our understanding of these mechanisms is incomplete with
open challenges and significant debate -- in part relating to the
translation of knowledge between scales, species, experimental
protocols, clinical cohorts, and individuals.

Perivascular pathways along the brain surface and within the brain
parenchyma have long been hypothesized to serve a designated role in
this context~\cite{rennels1985evidence, zhang1990interrelationships,
  ichimura1991distribution, carare2008solutes, iliff2012paravascular,
  foley2012realtime, hannocks2018molecular, van2024caa}. Recently, Eide and
Ringstad~\cite{eide2024functional} and Yamamoto et
al~\cite{yamamoto2024perivascular} demonstrated that perivascular
spaces (PVSs) define preferential pathways for molecular transport in
humans, with delayed periarterial enrichment in dementia
subtypes~\cite{eide2024functional}. Perivascular flow of cerebrospinal
fluid (CSF) clearly contributes to this transport, and is inherently
associated with vascular pulsations~\cite{hadaczek2006perivascular,
  iliff2013cerebral, bedussi2018paravascular, mestre2018flow,
  boster2023artificial, hirschler2024region}. Fluid mechanics
considerations point at intracranial pressure differences and shorter
wavelength vascular wall pulsations as drivers of directional net flow
and convection in the PVS~\cite{bilston2003arterial, rey2018pulsatile,
  daversin2020mechanisms, kedarasetti2020functional,
  gjerde2023directional, nozaleda2024arterial}, while longer waves
such as the pulse wave primarily contribute to oscillatory flow and
dispersion~\cite{asgari2016glymphatic, sharp2019dispersion, thomas2019fluid,
  kedarasetti2020arterial, troyetsky2021dispersion, 
  martinac2021phase}. In the bigger picture, CSF is produced by the
choroid plexus~\cite{damkier2013cerebrospinal,
  steffensen2018cotransporter, liu2020direct}, pulsates through the
ventricular system, cisterns, and subarachnoid space (SAS) in
synchrony with cardiac, respiratory, and neural
waves~\cite{greitz1993pulsatile, wagshul2011pulsating,
  sweetman2011cerebrospinal, fultz2019coupled, vinje2019respiratory,
  eide2021direction, causemann2022human, williams2023neural,
  zimmermann2023total}, and drains via the dural sinuses, meningeal
lymphatics, cranial nerves, or other efflux
pathways~\cite{proulx2021cerebrospinal}. However, how these
physiological factors and physical mechanisms integrate to enhance or
impair human intracranial molecular transport over larger spatial
scales and longer time scales remain unknown. A related key question
is to what extent surface PVSs are separated from the SAS by
structural barriers~\cite{zhang1990interrelationships,
  weller2005microscopic, bedussi2017paravascular,
  pizzo2018intrathecal, mestre2022periarteriolar,
  mollgard2023mesothelium, smets2024perivascular, eide2024functional},
and in turn to what extent such structural compartmentalization is a
prerequisite for effective perivascular transport.

% What are our research questions and how do we answer them
In this study, by leveraging geometric model reduction and
mixed-dimensional modelling~\cite{masri2024modelling}, structural
magnetic resonance (MR) images~\cite{hodneland2019new}, and high
performance computing, we introduce an integrated computational model
of intracranial molecular enrichment and clearance. Focusing on the
interplay between perivascular pathways, pulsatility and CSF flow
dynamics, the model predicts the temporal evolution and spatial
distribution of a solute concentration within a detailed geometric
representation of the human SAS and ventricular system, networks of
surface PVSs and the brain parenchyma. In terms of transport dynamics,
we account for heterogeneous diffusion, dispersive mixing induced by
cardiac and respiratory pulsatility in the CSF spaces and PVSs,
convective fluid flow driven by CSF production and peristaltic
pumping, as well as solute exchange and clearance across
semi-permeable membranes. Comparing with glymphatic MRI
studies~\cite{ringstad2017glymphatic, ringstad2018brain,
  watts2019measuring}, the in-silico predictions accurately represent
molecular enrichment patterns, timing and intercompartmental
distributions. This open platform~\cite{ZENODO} thus provides a
technological opportunity for qualitatively and quantitatively
exploring key open questions relating to molecular movement within the
human brain environment such as the role of pulsatility, perivascular
pathways, structural compartmentalization, or morphological
pathologies.

% Main findings and outlook
By exploring this high-dimensional parameter space, we propose that
the balance between CSF production and intracranial pulsatility is key
to shaping the large-scale features of intracranial enrichment
patterns, with the potential to span a wide range of individual and
cohort variability. Moreover, we predict that CSF production, cardiac-
and respiratory pulsatility is not sufficient to explain early
perivascular enrichment, but that fluid flow induced by low-frequency
vasomotion in surface periarterial spaces (on the order of 13 \textmu
m/s) is sufficient, even in the absence of a structural
compartmentalization of the PVS. Conversely, enlarged PVSs in the SAS
will cause a substantial reduction in cardiac- and vasomotion-driven
flow velocities, strongly delay perivascular transport, and thus
impair intracranial enrichment. These findings transfer, reconcile,
integrate and extend insights from clinical, experimental, and
theoretical studies, and lay a new foundation for in-silico studies of
personalized intrathecal drug delivery and brain clearance.

%% \begin{itemize}
%% \item
%%   Need better understanding of transport of solutes in intracranial spaces.
%% \item
%%   Need to integrate separate observations e.g. on diffusion in
%%   neuropil, PVS flow, net flow, effect of dispersion
%% \item
%%  Need to transfer and integrate insights from animal models,
%%   computational models and theoretical insights into the human case. 
%% \item
%%  Controversies need to be resolved.
%% \end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
%% \mer{MER: Target ~3100 words in 6 subsections}

%% \mer{MER: Results #1 is about 500 words}
\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure1.png}
     \caption{A) Illustration of the spatial domains of the model: CSF-filled space (ventricles and SAS) $\Omega_{\rm CSF}$, parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and the veins $\Lambda_{\rm vein}$ and their interfaces and boundaries (see Methods); B) 3D rendering of the actual geometry used in this study parenchyma, CSF (clipped, blue), arterial network (red), venous network (blue)); C) arterial network (red) and venous network (blue); D) illustration of the modelled clinical tracer injection procedure; E) time of simulated first tracer arrival after intrathecal injection (concentration exceeds $0.1\,$mol/l); F) time of first tracer arrival in the arterial PVS (top) and venous PVS (bottom); G) sagittal, coronal and axial view of computed tracer concentrations over T1-image 1, 6, 12 and 24\,h after injection (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); H) total amount of tracer in each cranial compartment over the first 24\,h; I) average tracer concentration in each compartment over the first 24\,h. }
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--C). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}
(\Cref{fig:results1}D). Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

Simulating a glymphatic MRI protocol~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2024functional}, we then predict the
spreading of $0.5$ mmol intrathecally injected Gadobutrol after its
appearance at the craniocervical junction
(\Cref{fig:results1}E--G). After one hour, the in-silico
tracer moves upwards in the SAS frontally of the brainstem, and
quickly reaches the supratentorial regions. Here, it spreads both
posterior through the quadrigeminal cistern and the longitudinal
fissure, and anteriorly through the outer SAS, reaching the top of the
cerebral cortex after around 12 hours. After 24 hours, the tracer
covers most of the brain surface (with the exception of some posterior
regions) and has penetrated several centimeters into the parenchymal
tissue. This pattern is reflected in the mean tracer concentrations in
each compartment, where the CSF space reaches its peak of
$1.4\,$mmol/l after 2 hours, followed by the arterial PVS
concentration peak ($2.9\,$mmol/l after 4 hours)
(\Cref{fig:results1}I). While the mean concentrations in these
compartments drop soon after peaking, the parenchymal tissue
slowly enriches with tracer over the first 24 hours, with a final
value of $0.3$ mmol/l. Overall, about 40\% of the total amount of
tracer remains in the cranium 24 hours post-injection, with the largest
share in the CSF (53\%), followed by the parenchyma (35\%), the
venous PVS (8\%) and the arterial PVS
(5\%) (\Cref{fig:results1}H).

% \mer{MER: Results #2 is about 600 words}
\subsection*{Reduced CSF pulsatility strongly shifts intracranial enrichment patterns}
\begin{figure}%[h!]
\centering 
\includegraphics[width = 0.95\textwidth]{figures/figure2.png}
\caption{\textbf{The balance between CSF pulsatility and production shapes molecular enrichment patterns.}
A) CSF spaces: the third and fourth ventricle (V3\&4), lateral ventricles (LV) and SAS; 
B) Boundaries and interfaces: the spinal subarachnoid space (SSAS), upper and lower arachnoid membrane (AM-U and AM-L), pia (Pia) and lateral ventricles (LV); 
C) Schematic of the CSF flow model; 
D) CSF flow induced by CSF production with outflow allowed across the
upper convexity (AM-U); 
}
\label{fig:csf}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
E) Histograms of the production-induced CSF velocities in the SAS and third and fourth ventricle (V3 \& V4) in terms of the relative frequency of the mean flow speed in each computational cell weighted by its volume; 
F) CSF flow induced at peak systolic blood inflow; 
G) Cardiac-induced dispersion factor $R_c$;
H) CSF flow induced at peak respiratory expansion; 
I) Respiration-induced dispersion factor $R_r$, $D = (1 + R_c + R_r) D^{\rm Gad}$; J--M): Histograms of the cardiac-induced flow velocities (J),
cardiac-induced dispersion factors (K),
respiratory flow velocities (L), respiratory dispersion factor (M);
N--Q): sagittal, coronal and axial planes of tracer concentration after 12h for different model variations: baseline (N), no CSF production (O), low dispersion (P), and high dispersion (Q).
}
\end{figure}

The enrichment patterns observed clinically after intrathecal
injection of contrast differ substantially between subjects and
between pathological conditions~\cite{ringstad2018brain,
  eide2021direction, eide2021impaired, eide2022altered}. These
neurological conditions are also associated with alterations in the
pulsatile flow of CSF in the ventricular system and
SAS~\cite{eide2021direction}. Clearly, key physiological factors such
as cardiac pulsatility and respiration easily differ between
individuals and between both pathological and physiological states. We
therefore next asked to what extent -- and how -- variations in CSF
pulsatility would affect the intracranial enrichment characteristics.
